Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-05-19 LICS/sections/appendix.tex
camera-ready and arxiv script

In this section we will define the category of semilattices that we use as a base for enrichment throughout the paper.

@@

In this section we will define the conditions under which the pushout in $\catname{EHyp}(\Sigma)$ exists, and we will explicitly construct such a pushout.
We will also show the uniqueness of a boundary pushout complement.
These condition will play a crucial role in showing that for each enrichment equation in $\textbf{PROP}^{+}(\Sigma)$ there is a sequence of rewrites in $\WellTypedMdaEcospans$.

@@

In this section we will elaborate on soundness of our translation $\llbracket - \rrbracket : \textbf{PROP}^{+}(\Sigma) \to \WellTypedMdaEcospans / \mathcal{S}$.

@@

\begin{proposition}[Proposition~\ref{prop:soundness} restatement]

@@

The category $\WellTypedMdaEcospans/{\mathcal{S}}$ is a semilattice-enriched PROP. 

@@

    To prove this statement we essentially need to show that the functor $\llbracket - \rrbracket$ is well-defined, \textit{i.e.}, if $f = g$ in $\textbf{PROP}^{+}(\Sigma)$ then $\llbracket f\rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ or $\llbracket f \rrbracket \Rrightarrow_{\mathcal{S}}^{*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans$.

@@

        \item $f = \mathcal{C}[l] = g$ for some morphism $l \in \textbf{PROP}^{+}(\Sigma)$.
        \item $f = \mathcal{C}[l]$ and $g = \mathcal{C}[r]$ where $l = r$ or $r = l$ is from $\mathcal{S}$.
        \item There is a sequence $w = (x_1, \ldots, x_n)$ such that $x_1 = f$ and $x_n = g$ and $x_i = x_{i+1}$ or $x_{i+1} = x_{i}$ in the sense above for $i < n$.

@@

        \item If $\mathcal{C}$ is empty, then either $f = g$ syntactically and $\llbracket f \rrbracket = \llbracket g \rrbracket$ or $f = l = r = g$ for some $l = r$ or $r = l$ in $\mathcal{S}$.
              The latter trivially yields the E-DPO diagram below

@@

                \textcolor{red}{Consider cases when $\mathcal{L}$ has no-input and no-output sub-hypergraph and when it does not. Basically, repeat the above reasoning here as well, but say that if $n+m$ is empty, then there is no pushout.}

@@

    The rest of the argument follows by induction on the length of $w$ and ultimately we have shown that if $f = g$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
    % there is a sequence of SMC and enrichment equations such that when applied to $f$ it is turned into $g$.
    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[]$ is defined by computing the pushout along the common interfaces.
    % We first note that any subterm of $g$ correspond to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence there is always a convex down-closed match of $\llbracket l \rrbracket$ within $\llbracket g \rrbracket$.
    % The mapping of external interfaces into $\llbracket l \rrbracket$ is mono and therefore the corresponding pushout complement exists.
    % We then need to show that for all $\llbracket l \rrbracket = \llbracket r \rrbracket$ in $\mathcal{S}$ the match satisfies the conditions in Proposition~\ref{prop:complement_existence} which will complete the first square of EDPOI and then check that the arrows in the second square satisfy the condition from Theorem~\ref{th:existence_of_pushouts}.

@@

\section{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:iso}
\begin{example}
    Consider cospans from Figure~\ref{fig:appendix:non-isomorphic-cospans}(a) where the $f_{int}, f_{ext}$ (respectively, $g_{int},g_{ext}$) maps are shown with red arrows.
    The left cospan represents a symmetry placed inside a hierarchical edge while the right one depicts a tensor product of two identities put inside the hierarchical edge.
    The colours of the vertices show the induced partition of the interfaces.
    We will try to build an isomorphism step-by-step.
    To make the diagram commute, clearly, it must be the case that $u_0 \mapsto u_0$ and $u_1 \mapsto u_1$.
    Then, because the isomorphism should preserve the order within the subset of green verices, it must be the case that $u_2 \mapsto u_2$ and $u_3 \mapsto u_3$.
    Similarly for the output interfaces.
    To further make the diagram commute, we need to map $v_2 \mapsto v_3$ and $v_3 \mapsto v_2$.
    This, however, contradicts the requirement that $g_{int};\gamma(w_2) = g_{int}(w_2)$.
    Hence, the two cospans are not isomorphic.
\end{example}

@@

\begin{figure*}
    \begin{subfigure}[T]{0.48\textwidth}
    \begin{subfigure}[T]{0.45\textwidth}
        \[
    \scalebox{0.6}{
        \tikzfig{../figures/appendix/sym_cospan}
    }    
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.6}{
                \tikzfig{../figures/appendix/id_x_id_cospan}
            }    
            \]
    \end{subfigure}
    \subcaption{\;}
\end{subfigure}
    \hfill
\begin{subfigure}[T]{0.48\textwidth}
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.6}{
                \tikzfig{../figures/appendix/id_x_id_cospan}
            }    
            \]
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.6}{
                \tikzfig{../figures/appendix/id_x_id_cospan_2}
            }    
            \]
    \end{subfigure}
    \subcaption{\;}
\end{subfigure}
    \caption{Non-isomorphic cospans (a) and isomorphic cospans (b)}
    \label{fig:appendix:non-isomorphic-cospans}
\end{figure*}

@@

Assume we want to perform a rewrite within the e-hypergraph in Figure~\ref{fig:dpo-stuck} (left).

@@

However, we are immediately stuck as there is no matching of the left-hand side of the rule within the graph as there is no pushout complement: the only way we can impose a constraint that an edge shall have a particular predecessor in the pushout is through its adjacent vertices, but the sub-hypergraph for $a;!$ does not have any as demonstrated in the right subfigure.

@@

We will start completing all the above DPO squares by first considering the structural rules.
We will include special unlabelled edges with no inputs and outputs to the interfaces to be able to impose the hierarchical relation on subgraphs with no inputs and outputs.
That is, the previously mentioned structural rule will become

@@

\section{Going from e-graph to e-hypergraph}

@@

\begin{definition}[Congruence invariant~\cite{EggPaper}]
Two e-nodes $f(a_{1}, \ldots, a_{n})$ and $f_{b_{1}, \ldots, b_{n}}$ are \textit{congruent}, if $find(a_{i}) = find(b_{i})$ for all $i$.
The equivalence relation on e-nodes must be closed under this congruence, that is, two congruent e-nodes must reside in the same e-class.
\end{definition}

@@

\begin{definition}
The pseudocode in Algorithm~\ref{alg:translation} defines the translation function $\llbracket \rrbracket : \catname{E}\text{-}\catname{graph} \to \WellTypedMdaEcospans$ that turns a \textit{canonical} (the one for which the above invariants hold) to a cospan of e-hypergraphs.
\end{definition}

@@

\begin{figure}
    \begin{subfigure}{0.4\linewidth}
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/e-graph-example_1}
}
\]
\end{subfigure}
\begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.6}{
    \tikzfig{../figures/appendix/e-hypergraph-example_1}
    }
    \]
\end{subfigure}
\end{figure}

@@

@@

>>>>

In this section we define the category of semilattices that we use as a base for enrichment throughout the paper.

@@

In this section we define the conditions under which the pushout in $\catname{EHyp}(\Sigma)$ exists, and we explicitly construct such a pushout.
We also show the uniqueness of a boundary pushout complement.
These condition will play a crucial role in showing that for each enrichment equation in $\textbf{PROP}^{+}(\Sigma)$ there is a sequence of rewrites in $\WellTypedMdaEcospans$.

@@

In this section we will elaborate on soundness and completeness of our translation $\llbracket - \rrbracket : \catname{S}^{+}(\Sigma, \mathcal{E}) \to \WellTypedMdaEcospans / \mathcal{S}, \mathcal{E}$.
First we recite the interpretation function $[]$ on $\Sigma$-terms as defined in~\cite{bonchi_string_2022-2} that induces a functor $\catname{S}(\Sigma) \to \MdaCospans$ and eventually a functor $\catname{S}^{+}(\Sigma) \to \MdaCospans^{+}$.
The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Figure~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneq [f] \otimes [g]$ and $[f;g] \Coloneq [f];[g]$ and by freeness induces the corresponding functor.

\begin{figure}
\[
\adjustbox{width=\textwidth}{
\tikzfig{../figures/appendix/interpretation}
}
\]
\caption{Base cases for $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$}
\label{fig:base_cases}
\end{figure}

\begin{lemma}[Lemma~\ref{lemma:cospans_plus_equiv} restatement]

    We have the following equivalence of $\catname{SLat}$-categories
    \[
    \MdaCospans^{+} \cong \MdaEcospans/\mathcal{S}~.
    \]
    \end{lemma}
    \begin{proof}
    We will denote the semilattice-enriched functor giving this equivalence as $M$.
    Observe that in $\MdaCospans^{+}$ we can consider only canonical morphisms of the form $f_{1} + \ldots + f_{n}$ where each of $f_{i}$ is an MDA cospan (in the sense of Definition~\ref{def:monogamy_hyp}) and for $i \not = j$, $f_{i} \not = f_{j}$.
    Then, $M$ is defined as follows.
    \begin{itemize}
    \item It maps each discrete hypergraph in $\MdaCospans^{+}$ to a corresponding discrete e-hypergraph in $\MdaEcospans/\mathcal{S}$.
    \item Every morphism $f_{1} + \ldots + f_{n}$ is mapped to $\text{In}(f_{1}) + \ldots + \text{In}(f_{n})$ where $\text{In}$ is the faithful functor induced by Remark~\ref{remark:embedding_functor}.
          This makes $M$ into $\catname{SLat}$-functor by construction.
    \end{itemize}
    To show the functor $M$ is faithful, suppose $f = f_{1} + \ldots + f_{n} \not = g_{1} + \ldots + g_{m} = g$.
    Then it must be that $Mf \not = Mg$.
    If $n \not = m$ then, immediately $Mf \not = Mg$ as $Mf$ and $Mg$ contain different number of non-isomorphic components.
    If $n = m$ then assume for a contradiction that $Mf = Mg$. 
    There exists a permutation $\sigma$, such that for every index $i$, $Mf_{i} = Mg_{\sigma(i)}$.
    Thus, by $\text{In}$ we have that $f_{i} = g_{\sigma(i)}$, implying $f = g$ which is a contradiction.
    Therefore, $Mf \not = Mg$ as required.

    To show that the functor is full, consider a morphism $f = f_{1} + \ldots + f_{n}$ in $\WellTypedMdaEcospans / \mathcal{S}$ (by Lemma~\ref{lemma:normal_form}). 
    As each of $f_{i}$ does not contain any hierarchical edges, they are in the image of $\text{In}$, which for each $f_{i}$ gives us $g_{i}$, such that $f_{i} = Mg_{i}$.
    This gives us a morphism $g = g_{1} + \ldots + {g_{n}}$ in $\MdaCospans^{+}$ such that $Mg = f$.
    This concludes that the functor is full.
    As it is also surjective on objects, it makes it into an equivalence.
\end{proof}

@@

\begin{theorem}[Theorem~\ref{thm:full-completeness} restatement]

@@

    Let $\mathcal{E}$ be a set of equations $l = r$ of $\Sigma$-terms, then we have the following equivalence of $\catname{SLat}$-categories
        \catname{S}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
    where $\mathcal{E}$ on the right is overloaded as a set of EDPOI rewrite rules containing $\langle \llbracket l \rrbracket, \llbracket r \rrbracket  \rangle$ and $\langle \llbracket r \rrbracket, \llbracket l \rrbracket  \rangle$ for every equation $l = r$.

@@

    The statement boils down to showing that the $\catname{SLat}$ functor defining the equivalence is well-defined, full and faithful as it is obviously surjective.
    Proof of fullness follows the same argument as in the case of $\llbracket - \rrbracket: \catname{S}^{+}(\Sigma) \to \WellTypedMdaEcospans / \mathcal{S}$, so we proceed with showing well-definedness and faithfulness.     
    The former is equivalent to showing that if $f = g$ in $\catname{S}^{+}(\Sigma)$ modulo $\mathcal{E}$ then $\llbracket f\rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S,E}~.$ or, equivalently, $\llbracket f \rrbracket \Rrightarrow_{\mathcal{E}}^{*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.

@@

        \item $f = \mathcal{C}[l] = g$ for some morphism $l \in \catname{S}^{+}(\Sigma)$;
        \item $f = \mathcal{C}[l]$ and $g = \mathcal{C}[r]$ for an equation $l = r$ or $r = l$ from $\mathcal{E}$;
        \item there is a sequence $w = (x_1, \ldots, x_n)$ such that $x_1 = f$ and $x_n = g$ and $x_i = x_{i+1}$ or $x_{i+1} = x_{i}$ in the sense above for $i < n$.

@@

        \item If $\mathcal{C}$ is empty, then either $f = g$ syntactically and $\llbracket f \rrbracket = \llbracket g \rrbracket$ or $f = l = r = g$ for some $l = r$ or $r = l$.
              The latter trivially yields the EDPO diagram below

@@

                % \textcolor{red}{Consider cases when $\mathcal{L}$ has no-input and no-output sub-hypergraph and when it does not. Basically, repeat the above reasoning here as well, but say that if $n+m$ is empty, then there is no pushout.}

@@

    The rest of the argument follows by induction on the length of $w$ and ultimately we have shown that if $f = g$ modulo $\mathcal{E}$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$  in $\WellTypedMdaEcospans / \mathcal{S}, \mathcal{E}$.

    Then, faithfulness amounts to showing that if $\llbracket f \rrbracket \Rrightarrow_{\llbracket \mathcal{E} \rrbracket}^{*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ then $f = g$ modulo $\mathcal{E}$ in $\catname{S}^{+}(\Sigma)$.
% By the theorem~\ref{thm:completeness_simple} if $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{S}$ then $f = g$, so we will show that the same holds when $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{E}$.
Consider a rewrite rule $\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle = \langle \mathcal{L}, \mathcal{R} \rangle$ such that $\llbracket f \rrbracket \Rrightarrow_{\langle \mathcal{L}, \mathcal{R} \rangle} \llbracket g \rrbracket$ and suppose
\[
	\llbracket f \rrbracket = n \xrightarrow{} n' \xrightarrow{} \mathcal{F} \xleftarrow{} m' \xleftarrow{} m ~.
\]
This implies the existence of a diagram as in~\ref{def:dpoi-e}.
We will proceed by induction on the level of the occurrence of $\mathcal{L}$ within $\mathcal{F}$.
\begin{itemize}
	\item Suppose the image of $\mathcal{L}$ in $\mathcal{F}$ is top-level.
		  Then, by lemma~\ref{lemma:decomposition}, there exists a decomposition of $\llbracket f \rrbracket$ in terms of MDA cospans
		\begin{align*}
			(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
			n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
			(i \xrightarrow{} i' \xrightarrow{} &\mathcal{L} \xleftarrow{} j' \xleftarrow{} j)
		\end{align*}
		By fullness of $\llbracket - \rrbracket$ (from Theorem~\ref{thm:completeness_simple}) there exist $c_{1}$ and $c_{2}$ such that 
		\[
		\llbracket c_{1} \rrbracket = n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k \qquad \llbracket c_{2} \rrbracket = j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m
		\]
		By functoriality, $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes l);c_{2} \rrbracket$ and by faithfulness we have that $f = c_{1};(id \otimes l);c_{2}$ to which we can apply $\langle l, r \rangle$ and get $g' = c_{1};(id \otimes r);c_{2}$.
		We then have $\llbracket g' \rrbracket = \llbracket c_{1};(id \otimes r);c_{2} \rrbracket =  \llbracket g \rrbracket$ since the boundary complement is unique (Proposition ~\ref{prop:boundary_unique}) and $g' = g$ by faithfulness.

		\item If the image of $\mathcal{L}$ is not top-level, then either $\llbracket f \rrbracket$ is decomposable as
		\begin{align*}
			(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
			n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
			(i \xrightarrow{} i' \xrightarrow{} &\mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)
		\end{align*}
		such that $\mathcal{L}$ is a convex down-closed sub-e-hypergraph of $F'$.
		Since the image of $\mathcal{L}$ lies within $\mathcal{F}'$, there is an EDPO rewrite step that turns
		\[
			(i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)	
		\]
		into
		\[
			(i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)	
		\]
		via $\langle \mathcal{L}, \mathcal{R} \rangle$.
		By fullness and inductive hypothesis, there exist $f'$ and $g'$ such that 
		\[
			\llbracket f' \rrbracket = i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j \qquad \llbracket g' \rrbracket = (i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)
		\]
		and that $f' = g'$.
		By decomposition and fullness we have $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes f');c_{2} \rrbracket$ and $f = c_{1};(id \otimes f');c_{2}$ by faithfulness.
		By applying $f' = g'$ we get $f = g'' = c_{1};(id \otimes g');c_{2}$.
		Finally, $\llbracket g'' \rrbracket = \llbracket c_{1};(id \otimes g');c_{2} \rrbracket = \llbracket g \rrbracket$ (by the uniqueness of the boundary complement) and $g'' = g = f$ by faithfulness.

		\item or, $\llbracket f \rrbracket$ is decomposable as
		\begin{align*}
			n_1 \xrightarrow{} n_1' \xrightarrow{} &\mathcal{F}_1 \xleftarrow{} m_1' \xleftarrow{} m_1\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_p \xrightarrow{} n_p' \xrightarrow{} &\mathcal{F}_p \xleftarrow{} m_p' \xleftarrow{} m_p\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_k \xrightarrow{} n_k' \xrightarrow{} &\mathcal{F}_k \xleftarrow{} m_k' \xleftarrow{} m_k
		\end{align*}
		such that the image of $\mathcal{L}$ lies within $\mathcal{F}_{p}$.
		By inductive hypothesis we have that $\llbracket f_{p} \rrbracket = n_p \xrightarrow{} n_{p}' \xrightarrow{} \mathcal{F}_{p} \xleftarrow{} m_{p}' \xleftarrow{} m_{p}$ rewrites into $\llbracket g_{p} \rrbracket = n_p \xrightarrow{} n_{p}'' \xrightarrow{} \mathcal{G}_{p} \xleftarrow{} m_{p}'' \xleftarrow{} m_{p}$ such that $f_{p} = g_{p}$.
		By fullness and the fact that $\llbracket - \rrbracket$ is enriched, there exist $\llbracket f_1 + \ldots + f_{p} + \ldots + f_{k} \rrbracket = \llbracket f_{1} \rrbracket + \ldots + \llbracket f_{p} \rrbracket + \ldots + \llbracket f_{k} \rrbracket = \llbracket f \rrbracket$.
		By applying $f_{p} = g_{p}$ we get $f = g''$ and $\llbracket g'' \rrbracket = \llbracket f_{1} + \ldots + g_{p} + \ldots f_{k} \rrbracket = \llbracket g \rrbracket$ and $f = g'' = g$.
\end{itemize}
    % there is a sequence of SMC and enrichment equations such that when applied to $f$ it is turned into $g$.
    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[]$ is defined by computing the pushout along the common interfaces.
    % We first note that any subterm of $g$ correspond to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence there is always a convex down-closed match of $\llbracket l \rrbracket$ within $\llbracket g \rrbracket$.
    % The mapping of external interfaces into $\llbracket l \rrbracket$ is mono and therefore the corresponding pushout complement exists.
    % We then need to show that for all $\llbracket l \rrbracket = \llbracket r \rrbracket$ in $\mathcal{S}$ the match satisfies the conditions in Proposition~\ref{prop:complement_existence} which will complete the first square of EDPOI and then check that the arrows in the second square satisfy the condition from Theorem~\ref{th:existence_of_pushouts}.

@@

\section{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:iso}

First we will explore the notion of isomorphism of two extended cospans in more detail.
Our example will be Figure~\ref{fig:appendix:non-isomorphic-cospans}.
Technically, we do not allow hierarchical edges with no delimiting dashed line, but here, for the purposes of illustration we make an exception to make the example compact.
\begin{example}
    Consider cospans from Figure~\ref{fig:appendix:non-isomorphic-cospans}(a) where the $f_{int}, f_{ext}$ (respectively, $g_{int},g_{ext}$) maps are shown with red arrows.
    The left cospan represents a symmetry placed inside a hierarchical edge while the right one depicts a tensor product of two identities put inside the hierarchical edge.
    The colours of the vertices show the induced partition of the interfaces.
    We will try to build an isomorphism step-by-step.
    To make the diagram commute, clearly, it must be the case that $u_0 \mapsto u_0$ and $u_1 \mapsto u_1$.
    Then, because the isomorphism should preserve the order within the subset of green vertices, it must be the case that $u_2 \mapsto u_2$ and $u_3 \mapsto u_3$.
    Similarly for the output interfaces.
    To further make the diagram commute, we need to map $v_2 \mapsto v_3$ and $v_3 \mapsto v_2$.
    This, however, contradicts the requirement that $g_{int};\gamma(w_2) = g_{int}(w_2)$.
    Hence, the two cospans are not isomorphic.
\end{example}

@@

Then we will elaborate on the notion of a boundary complement.

\begin{example}

Consider a pushout square in Figure~\ref{fig:not_boundary_complement} (a) where the labels suggest how arrows map vertices both from the interfaces and between the e-hypergraphs.
While this is a valid pushout square, such pushout complement does not correspond to a rewrite of $\Sigma^{+}$-terms --- as there is no redex for $f \otimes g$ within $(f + g);g$ even modulo SMC equations.
This pushout fails to be a boundary complement by violating conditions (3) and (4) of Definition~\ref{def:boundary_new}.
Now consider the square in subfigure (b): it fails to be a boundary complement by violating conditions (4) and (7).
\end{example}

\begin{example}

Now consider a full DPO square in Figure~\ref{fig:not_boundary_complement} (c).
First note that the pushout complement is a boundary complement as it satisfies all the conditions in Definition~\ref{def:boundary_new}.
In particular, observe that in order to make an extended MDA cospan with the boundary complement as a carrier, strict internal interfaces should be empty.
Indeed, strict internal interfaces of the left-hand side coincide with the ones of the source e-hypergraph.
Then, for the resulting e-hypergraph to form a valid well-typed MDA extended cospan we need to introduce the strict internal interfaces of the right-hand side of the rewrite rule.
\end{example}

\begin{figure*}
    \begin{subfigure}[T]{0.48\textwidth}
    \begin{subfigure}[T]{0.45\textwidth}
        \[
    \scalebox{0.55}{
        \tikzfig{../figures/appendix/sym_cospan}
    }    
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.55}{
                \tikzfig{../figures/appendix/id_x_id_cospan}
            }    
            \]
    \end{subfigure}
    \subcaption{\;}
\end{subfigure}
    \hfill
\begin{subfigure}[T]{0.48\textwidth}
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.55}{
                \tikzfig{../figures/appendix/id_x_id_cospan}
            }    
            \]
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.45\textwidth}
        \[
            \scalebox{0.55}{
                \tikzfig{../figures/appendix/id_x_id_cospan_2}
            }    
            \]
    \end{subfigure}
    \subcaption{\;}
\end{subfigure}
    \caption{Non-isomorphic cospans (a) and isomorphic cospans (b)}
    \label{fig:appendix:non-isomorphic-cospans}
\end{figure*}

\begin{figure}
    \begin{subfigure}{0.48\textwidth}
\[
\adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/boundary_complement_non_example}
}
\]
\subcaption{Boundary complement non-example}
    \end{subfigure}
\hfill
    \begin{subfigure}{0.48\textwidth}
        \[
        \adjustbox{scale=0.55}{
            \tikzfig{../figures/appendix/boundary_complement_non_example_2}
        }
        \]
            \subcaption{Boundary complement non-example}
            \end{subfigure}
\begin{subfigure}{\textwidth}
    \[
    \adjustbox{scale=0.55}{
        \tikzfig{../figures/appendix/boundary_complement_example}
    }
    \]
        \subcaption{Complete DPO square}
        \end{subfigure}
\caption{DPO examples}
\label{fig:not_boundary_complement}
\end{figure}

@@

Assume we want to perform a rewrite within the e-hypergraph in Figure~\ref{fig:dpo-stuck} (left).
However, we are immediately stuck as there is no matching of the left-hand side of the rule within the graph as there is no pushout complement: the only way we can impose a constraint that an edge shall have a particular predecessor in the pushout is through its adjacent vertices, but the sub-hypergraph for $a;!$ does not have any as demonstrated in the right subfigure.

@@

@@

We will start completing all the above DPO squares by first considering the structural rules.
We will include special unlabelled edges with no inputs and outputs to the interfaces to be able to impose the hierarchical relation on subgraphs with no inputs and outputs and will not impose any restrictions on labelling preservation when mapping these special edges.
That is, the previously mentioned structural rule will become

@@

The Definition~\ref{def:boundary_new} will then need a few additional constraints.
\begin{itemize}
\item For all top-level edges $e_{i}, e_{j}$ of type $0 \to 0$ in $\mathcal{L}$ it must be either $[m(e_{i})) = [m(e_{j})) = \varnothing$ or $m(e_{i}) \consistency m(e_{j})$.
\item For all edges $e_{i}, e_{j}$ of type $0 \to 0$ in $i + j$ it must be either $[[c_1,c_2](e_{i})) = [[c_1,c_2](e_{j})) = \varnothing$ or $[c_1,c_2](e_{i}) \consistency [c_1,c_2](e_{j})$.
\item For all edges $e$ of type $0 \to 0$ in $i + j$ it must be that $l([c_1,c_2](e)) = l(e) = \bot$ and $[c_1,c_2](e)$ must be maximal.
\end{itemize}
Then, when constructing the corresponding extended cospans, such auxiliary edges must be ignored in the interfaces.

\section{E-graph interpretation in $\catname{MEHypI}(\Sigma_{C}) / \mathcal{S}, \mathcal{E}_{C}$}
\label{sec:appendix:e-graph-translation}

In this section we will elaborate on the interpretation of e-graphs as morphisms of $\catname{MEHypI}(\Sigma_{C}) / \mathcal{S}, \mathcal{E}_{C}$.
We will first recite the formal definition of an e-graph.

@@

\begin{definition}[Congruence invariant~\cite{EggPaper}]
Two e-nodes $f(a_{1}, \ldots, a_{n})$ and $f_{b_{1}, \ldots, b_{n}}$ are \textit{congruent}, if $\text{find}(a_{i}) = \text{find}(b_{i})$ for all $i$.
The equivalence relation on e-nodes must be closed under this congruence, that is, two congruent e-nodes must reside in the same e-class.
\end{definition}

@@

\begin{definition}
The pseudocode in Algorithm~\ref{alg:translation} defines the translation function $\llbracket \rrbracket : \catname{E}\text{-}\catname{graph} \to \catname{MEHypI}(\Sigma_{C}) / \mathcal{S}, \mathcal{E}_{C}$ that turns a \textit{canonical} (the one for which the above invariants hold) \textit{acyclic} \textit{connected} e-graph into a cospan of e-hypergraphs.
\end{definition}

@@

\begin{figure}
    \begin{subfigure}[C]{0.4\linewidth}
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/e-graph-example_1}
}
\]
\caption{\;}
\end{subfigure}
\hfill
\begin{subfigure}[C]{0.4\linewidth}
    \[
    \adjustbox{scale=0.6}{
    \tikzfig{../figures/appendix/e-hypergraph-example_1}
    }
    \]
\caption{\;}
\end{subfigure}
\caption{E-graph (a) and a corresponding morphism of $\catname{MEHypI}(\Sigma \cup \Sigma_{C})$ (b)}
\label{fig:e-graph-to-cospan}
\end{figure}

@@

\begin{remark}
Technically, $\mathcal{E}_{C}$ is not induced by equations between $\Sigma$-terms, but rather between $\Sigma^{+}$-terms as we can apply naturality of copy and delete to boxes as well.
This does not pose any issues as the proof of Theorem~\ref{thm:full-completeness} works for any $\mathcal{E}$ as long as the conditions of Definition~\ref{def:boundary_new} are satisfied.
\end{remark}

@@

\section{Additional figures}

Additional translation steps

\begin{figure}
    \[
    \adjustbox{width=\linewidth}{
    \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-c-e}
    }
    \]
    \caption{Translation (c)-(d)}
    \end{figure}
\begin{figure}

\[
\adjustbox{width=\linewidth}{
\tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-d-e}
}
\]
\caption{Translation (d)-(e)}
\end{figure}
